---
title: "Part 1 — Data Serialization in R"
title-long: "Fast & Big Data in R: Storing, Moving, and Benchmarking Data on Disk"
date: "2026-06-DD"   # TODO: confirm seminar date
description: "How R reads and writes data — RDS, qs2, fst, CSV, Feather, Parquet, and DuckDB — with benchmarks across 1k to 1M rows and practical recommendations for development-data workflows."
author: "Eduard Bukin"   # TODO: confirm speaker
categories: [R, serialization, qs2, fst, arrow, parquet, duckdb, benchmarks]
code-tools: true
code-fold: false
---

## What is Serialization?

::: {.unnumbered .unlisted style="font-family: 'Times New Roman', serif; font-size: 0.85em;"}

> Do not expect to write data to a file (such as with `write.table`), read the
> data back into R and have that be precisely the same as the original. That is
> doing two translations, and there is often something lost in translation.
>
> — Burns, P. (2011). *The R Inferno* (§ 8.3.10, p 105). burns-stat.com.

:::

Serialization converts an R object into a byte stream for storage or transmission, then reconstructs it exactly. Unlike text formats (CSV), which convert data to characters and discard type information, binary serialization preserves the complete object: column types, factor levels, class attributes, and custom metadata.

## R's Serialization Toolkit

Seven approaches cover most use cases, grouped by storage strategy.

### Binary R-native

Store any R object as bytes; reconstruct exactly with no post-processing.

| Package | Write / Read | Notes |
|---|---|---|
| base R | `saveRDS()` / `readRDS()` | Any R object; gzip/bzip2/xz compression; single-threaded |
| qs2 | `qs_save()` / `qs_read()` | Any R object; zstd/lz4; multi-threaded — 5–10× faster than saveRDS |
| fst | `write_fst()` / `read_fst()` | Data frames only; LZ4/ZSTD; multi-threaded; random column/row access |

### Text-based

Convert data to human-readable text; type information is lost and must be inferred on read.

| Package | Write / Read | Notes |
|---|---|---|
| base R | `write.csv()` / `read.csv()` | Portable; slow; single-threaded |
| data.table | `fwrite()` / `fread()` | Same format; 10–100× faster; smart type inference |

### Cross-platform binary

Use C++ libraries (Apache Arrow, DuckDB) to access R memory directly — zero-copy where possible, language-agnostic output.

| Package | Write / Read | Notes |
|---|---|---|
| arrow | `write_feather()` / `read_feather()` | Arrow columnar format; fastest I/O; largest files |
| arrow | `write_parquet()` / `read_parquet()` | Columnar with compression; column/row selection; cloud-native |
| duckdb | `dbWriteTable()` / `dbReadTable()` | Embedded SQL database; larger-than-RAM support |

## Native R Serialization in Practice

### Basic usage

```{r}
#| label: basic-serialization
#| results: hold
df <- data.frame(
  id    = 1:3,
  name  = c("Alice", "Bob", "Charlie"),
  score = c(95.5, 87.3, 92.1)
)
saveRDS(df, "temp_df.rds")
df_restored <- readRDS("temp_df.rds")
identical(df, df_restored)
```

### Format and compression options

```{r}
#| label: format-and-compression
#| results: hold

# Format version 3 supports ALTREP (e.g. 1:1000000 stored as a range, not all ints)
saveRDS(df, "temp_v2.rds", version = 2)
saveRDS(df, "temp_v3.rds", version = 3)
cat("Version 2:", file.size("temp_v2.rds"), "bytes\n")
cat("Version 3:", file.size("temp_v3.rds"), "bytes\n")

# Compression: FALSE = fastest write; "xz" = smallest file
large_df <- data.frame(x = rep(1:100, 100), y = rnorm(10000),
                       z = sample(letters, 10000, replace = TRUE))
saveRDS(large_df, "temp_none.rds",  compress = FALSE)
saveRDS(large_df, "temp_gzip.rds",  compress = "gzip")
saveRDS(large_df, "temp_bzip2.rds", compress = "bzip2")
saveRDS(large_df, "temp_xz.rds",   compress = "xz")
cat("No compression:", file.size("temp_none.rds"),  "bytes\n")
cat("gzip:          ", file.size("temp_gzip.rds"),  "bytes\n")
cat("bzip2:         ", file.size("temp_bzip2.rds"), "bytes\n")
cat("xz:            ", file.size("temp_xz.rds"),   "bytes\n")
```

### Binary vs text encoding

```{r}
#| label: binary-vs-text
#| results: hold
raw_bytes   <- serialize(df, connection = NULL)
ascii_bytes <- serialize(df, connection = NULL, ascii = TRUE)
head(raw_bytes,   16)
head(ascii_bytes, 16)
cat("Binary:", length(raw_bytes),   "bytes\n")
cat("ASCII: ", length(ascii_bytes), "bytes\n")
```

### Attributes and types are preserved

```{r}
#| label: attributes-and-types
#| results: hold

# Custom attributes and subclasses survive round-trips
df_special <- df
attr(df_special, "created_date") <- Sys.Date()
attr(df_special, "source") <- "example data"
class(df_special) <- c("my_special_df", "data.frame")
saveRDS(df_special, "temp_special.rds")
attributes(readRDS("temp_special.rds"))

# All column types are preserved exactly
df_types <- data.frame(
  int_col  = 1:3,
  dbl_col  = c(1.1, 2.2, 3.3),
  chr_col  = c("a", "b", "c"),
  fct_col  = factor(c("low", "high", "medium")),
  date_col = as.Date(c("2026-01-01", "2026-01-02", "2026-01-03")),
  stringsAsFactors = FALSE
)
saveRDS(df_types, "temp_types.rds")
str(df_types)
str(readRDS("temp_types.rds"))
```

```{r}
#| label: cleanup-temp-files
#| include: false
file.remove(c("temp_df.rds", "temp_v2.rds", "temp_v3.rds",
              "temp_none.rds", "temp_gzip.rds", "temp_bzip2.rds", "temp_xz.rds",
              "temp_special.rds", "temp_types.rds"))
```

## Format Deep Dives

### qs2 — fast serialization of any R object

The `qs2` package stores any R object (data frames, models, lists, environments) with
multi-threaded zstd/lz4 compression — typically 5–10× faster than `saveRDS()`.

```{r}
#| label: qs2-demo
#| eval: false
library(qs2)

qs_save(large_data, "data.qs2")                  # uses all available cores
qs_save(large_data, "data.qs2", nthreads = 4)    # or specify explicitly
qs2::qs_threads()                                 # check active thread count

# Save a complete analysis workspace — models, predictions, fitted objects
analysis_state <- list(
  model       = trained_rf_model,
  predictions = pred_results,
  metrics     = performance_metrics,
  metadata    = list(timestamp = Sys.time(), r_version = R.version.string)
)
qs_save(analysis_state, "analysis_cache.qs2")
analysis_state <- qs_read("analysis_cache.qs2")   # restore in one call
```

### fst — columnar binary with random access

`fst` stores data frames in a columnar format that supports partial reads without a
full scan — useful when working with wide data frames or large files where you
only need a subset of columns or a row range.

```{r}
#| label: fst-demo
#| eval: false
library(fst)

write_fst(df, "data.fst", compress = 50)    # 0 = fastest write; 100 = smallest file

df_full    <- read_fst("data.fst")                                         # everything
df_cols    <- read_fst("data.fst", columns = c("region", "income"))        # columns only
df_rows    <- read_fst("data.fst", from = 1000, to = 5000)                 # row range
df_partial <- read_fst("data.fst", columns = c("region", "income"),
                       from = 1000, to = 5000)                             # both
```

Compression and I/O happen on background threads, so `fst` can write faster than disk
speed — compression and disk writes overlap.

### Arrow Parquet — columnar analytics format

Parquet is the standard format for analytical workloads: compressed, columnar, and
readable by Python, Spark, BigQuery, DuckDB, and many other tools without conversion.

```{r}
#| label: parquet-demo
#| eval: false
library(arrow)
library(dplyr)

# Write and read
write_parquet(data_100k, "survey.parquet")
read_parquet("survey.parquet")
read_parquet("survey.parquet", col_select = c("id", "region", "income"))

# Compression options
write_parquet(data_100k, "fast.parquet",       compression = "snappy")     # faster
write_parquet(data_100k, "small.parquet",      compression = "zstd",
              compression_level = 9)                                        # smaller

# Partitioned dataset — splits into region=/year= directory tree
write_dataset(data_100k, path = "survey_parts", partitioning = c("region", "year"))
ds <- open_dataset("survey_parts")

# Filter + aggregate without loading the full file — only matching partitions are read
ds |>
  filter(region == "A", year >= 2023) |>
  group_by(country) |>
  summarise(avg_income = mean(income), n = n()) |>
  collect()
```

Replace the local path with `"s3://bucket/prefix/"` or an HTTPS URL and the same
code reads from cloud storage — only the requested columns and row groups are
downloaded.

### DuckDB — embedded analytical database

DuckDB runs as an in-process library (no server), speaks SQL and dplyr, reads Parquet
directly, and spills to disk automatically when data exceeds RAM.

```{r}
#| label: duckdb-demo
#| eval: false
library(duckdb)
library(dplyr)

con <- dbConnect(duckdb())              # in-memory
con <- dbConnect(duckdb(), "data.duckdb")  # persistent file

# Write R data frame into DuckDB
dbWriteTable(con, "survey", data_100k)

# SQL query
dbGetQuery(con, "
  SELECT region, AVG(income) AS avg_income, COUNT(*) AS n
  FROM survey WHERE employed = TRUE
  GROUP BY region ORDER BY avg_income DESC
")

# Same query via dplyr
tbl(con, "survey") |>
  filter(employed) |>
  group_by(region) |>
  summarise(avg_income = mean(income), n = n()) |>
  arrange(desc(avg_income)) |>
  collect()

# Query Parquet directly — no import step
dbGetQuery(con, "
  SELECT region, year, AVG(income) AS avg_income
  FROM read_parquet('survey_parts/**/*.parquet')
  WHERE year >= 2023
  GROUP BY region, year
")

dbDisconnect(con, shutdown = TRUE)
```

## Benchmarking

```{r}
#| label: generate-benchmark-data
#| message: false

library(tidyverse)
library(data.table)
library(fst)
library(qs2)
library(arrow)
library(duckdb)

set.seed(7892)

generate_test_data <- function(n_rows) {
  data.frame(
    id          = 1:n_rows,
    region      = factor(sample(LETTERS[1:5], n_rows, replace = TRUE)),
    country     = sample(c("Kenya", "Uganda", "Tanzania", "Rwanda", "Burundi",
                           "Ethiopia", "Somalia", "Sudan", "Chad", "Niger",
                           "Nigeria", "Ghana", "Senegal", "Mali", "Burkina Faso",
                           "Benin", "Togo", "Cameroon", "Congo", "DRC"),
                         n_rows, replace = TRUE),
    year        = sample(2020:2025, n_rows, replace = TRUE),
    income      = rlnorm(n_rows, meanlog = 8, sdlog = 1.5),
    employed    = sample(c(TRUE, FALSE), n_rows, replace = TRUE, prob = c(0.6, 0.4)),
    survey_date = as.Date("2020-01-01") + sample(0:2000, n_rows, replace = TRUE),
    notes       = sample(c("Complete", "Partial", "Missing data", NA, "Verified",
                           "Pending review", "Approved"), n_rows, replace = TRUE)
  )
}

data_1k   <- generate_test_data(1e3)
data_10k  <- generate_test_data(1e4)
data_100k <- generate_test_data(1e5)
data_1m   <- generate_test_data(1e6)

data.frame(
  Dataset       = c("1k rows", "10k rows", "100k rows", "1M rows"),
  `Size in RAM` = c(
    format(object.size(data_1k),   units = "MB"),
    format(object.size(data_10k),  units = "MB"),
    format(object.size(data_100k), units = "MB"),
    format(object.size(data_1m),   units = "MB")
  ),
  check.names = FALSE
) |> knitr::kable(caption = "Dataset sizes in memory")
```

The **Write/Read RAM (MB)** columns report the additional R-heap memory allocated
at the peak of a single operation, measured with `gc()`. Arrow (feather/parquet)
and DuckDB manage large allocations in C++ off the R heap, so their R-side figures
understate true peak usage.

```{r}
#| label: benchmark-helpers
#| include: false

mem_of <- function(expr) {
  g0 <- gc(reset = TRUE, full = TRUE)
  before <- sum(g0[, 2])
  force(expr)
  g1 <- gc(full = TRUE)
  round(sum(g1[, 6]) - before, 1)
}

bench_write_read <- function(data, write_fn, read_fn, filename, n_rep = 10) {
  fpath <- file.path(bench_dir, filename)
  write_times <- replicate(n_rep, {
    start <- Sys.time()
    write_fn(data, fpath)
    as.numeric(difftime(Sys.time(), start, units = "secs")) * 1000
  })
  read_times <- replicate(n_rep, {
    start <- Sys.time()
    result <- read_fn(fpath)
    as.numeric(difftime(Sys.time(), start, units = "secs")) * 1000
  })
  write_mem <- mem_of(write_fn(data, fpath))
  read_mem  <- mem_of(read_fn(fpath))
  fsize <- if (file.exists(fpath)) round(file.size(fpath) / 1024^2, 2) else 0
  list(
    write_ms     = round(median(write_times), 1),
    read_ms      = round(median(read_times), 1),
    size_mb      = fsize,
    write_mem_mb = write_mem,
    read_mem_mb  = read_mem
  )
}

bench_duckdb <- function(data, n_rep = 10) {
  fpath_db <- file.path(bench_dir, "test.duckdb")
  con <- dbConnect(duckdb::duckdb(), fpath_db)
  write_times_db <- replicate(n_rep, {
    start <- Sys.time()
    dbWriteTable(con, "data", data, overwrite = TRUE)
    as.numeric(difftime(Sys.time(), start, units = "secs")) * 1000
  })
  read_times_db <- replicate(n_rep, {
    start <- Sys.time()
    result <- dbReadTable(con, "data")
    as.numeric(difftime(Sys.time(), start, units = "secs")) * 1000
  })
  write_mem_db <- mem_of(dbWriteTable(con, "data", data, overwrite = TRUE))
  read_mem_db  <- mem_of(dbReadTable(con, "data"))
  fsize <- round(file.size(fpath_db) / 1024^2, 2)
  dbDisconnect(con, shutdown = TRUE)
  list(
    write_ms     = round(median(write_times_db), 1),
    read_ms      = round(median(read_times_db), 1),
    size_mb      = fsize,
    write_mem_mb = write_mem_db,
    read_mem_mb  = read_mem_db
  )
}

results_table <- function(results) {
  data.frame(
    Method           = names(results),
    `Write (ms)`     = sapply(results, function(x) x$write_ms),
    `Write RAM (MB)` = sapply(results, function(x) x$write_mem_mb),
    `Read (ms)`      = sapply(results, function(x) x$read_ms),
    `Read RAM (MB)`  = sapply(results, function(x) x$read_mem_mb),
    `File Size (MB)` = sapply(results, function(x) x$size_mb),
    check.names = FALSE,
    row.names = NULL
  )
}
```

::: panel-tabset

## 1M Rows

```{r}
#| label: benchmark-1m
#| cache: true
#| message: false
#| warning: false

bench_dir <- tempfile(); dir.create(bench_dir)

results_1m <- list(
  saveRDS    = bench_write_read(data_1m, function(d,f) saveRDS(d,f), readRDS,     "test.rds",     n_rep = 5),
  qs2        = bench_write_read(data_1m, qs_save,                    qs_read,      "test.qs2",     n_rep = 5),
  fst        = bench_write_read(data_1m, write_fst,                  read_fst,     "test.fst",     n_rep = 5),
  data.table = bench_write_read(data_1m, fwrite,                     fread,        "test.csv",     n_rep = 5),
  feather    = bench_write_read(data_1m, write_feather,              read_feather, "test.feather", n_rep = 5),
  parquet    = bench_write_read(data_1m, write_parquet,              read_parquet, "test.parquet", n_rep = 5)
)
results_1m$duckdb <- bench_duckdb(data_1m, n_rep = 5)

knitr::kable(results_table(results_1m), caption = "Performance comparison — 1M rows")
unlink(bench_dir, recursive = TRUE)
```

## 100k Rows

```{r}
#| label: benchmark-100k
#| cache: true
#| message: false
#| warning: false

bench_dir <- tempfile(); dir.create(bench_dir)

results_100k <- list(
  saveRDS    = bench_write_read(data_100k, function(d,f) saveRDS(d,f), readRDS,     "test.rds"),
  qs2        = bench_write_read(data_100k, qs_save,                    qs_read,      "test.qs2"),
  fst        = bench_write_read(data_100k, write_fst,                  read_fst,     "test.fst"),
  data.table = bench_write_read(data_100k, fwrite,                     fread,        "test.csv"),
  feather    = bench_write_read(data_100k, write_feather,              read_feather, "test.feather"),
  parquet    = bench_write_read(data_100k, write_parquet,              read_parquet, "test.parquet")
)
results_100k$duckdb <- bench_duckdb(data_100k)

knitr::kable(results_table(results_100k), caption = "Performance comparison — 100k rows")
unlink(bench_dir, recursive = TRUE)
```

## 10k Rows

```{r}
#| label: benchmark-10k
#| cache: true
#| message: false
#| warning: false

bench_dir <- tempfile(); dir.create(bench_dir)

results_10k <- list(
  saveRDS    = bench_write_read(data_10k, function(d,f) saveRDS(d,f), readRDS,     "test.rds"),
  qs2        = bench_write_read(data_10k, qs_save,                    qs_read,      "test.qs2"),
  fst        = bench_write_read(data_10k, write_fst,                  read_fst,     "test.fst"),
  data.table = bench_write_read(data_10k, fwrite,                     fread,        "test.csv"),
  feather    = bench_write_read(data_10k, write_feather,              read_feather, "test.feather"),
  parquet    = bench_write_read(data_10k, write_parquet,              read_parquet, "test.parquet")
)
results_10k$duckdb <- bench_duckdb(data_10k)

knitr::kable(results_table(results_10k), caption = "Performance comparison — 10k rows")
unlink(bench_dir, recursive = TRUE)
```

## 1k Rows

```{r}
#| label: benchmark-1k
#| cache: true
#| message: false
#| warning: false

bench_dir <- tempfile(); dir.create(bench_dir)

results_1k <- list(
  saveRDS    = bench_write_read(data_1k, function(d,f) saveRDS(d,f), readRDS,     "test.rds"),
  qs2        = bench_write_read(data_1k, qs_save,                    qs_read,      "test.qs2"),
  fst        = bench_write_read(data_1k, write_fst,                  read_fst,     "test.fst"),
  data.table = bench_write_read(data_1k, fwrite,                     fread,        "test.csv"),
  feather    = bench_write_read(data_1k, write_feather,              read_feather, "test.feather"),
  parquet    = bench_write_read(data_1k, write_parquet,              read_parquet, "test.parquet")
)
results_1k$duckdb <- bench_duckdb(data_1k)

knitr::kable(results_table(results_1k), caption = "Performance comparison — 1k rows")
unlink(bench_dir, recursive = TRUE)
```

:::

## Summary & Recommendations

Choose based on your primary use case. Rows are properties; columns are tools.

| | saveRDS | qs2 | fst | fwrite / fread | feather | parquet | DuckDB |
|:---|:---:|:---:|:---:|:---:|:---:|:---:|:---:|
| **Write speed** | slow | fast | fast | fast | fastest | moderate | moderate |
| **Read speed** | slow | fast | fastest | fast | fastest | fast | moderate |
| **File size** | small | small | medium | largest | large | smallest | medium |
| **Compressed** | yes | yes | optional | no | no | yes | yes |
| **Multi-threaded** | no | yes | yes | yes | yes | yes | yes |
| **Stores any R object** | ✓ | ✓ | df only | df only | df only | df only | df only |
| **Select columns on read** | – | – | ✓ | – | ✓ | ✓ | ✓ |
| **Select rows on read** | – | – | range only | – | – | ✓ | ✓ SQL |
| **Cross-language** | R only | R only | R only | any text tool | Arrow | Python/Spark/etc. | any |
| **Larger-than-RAM** | – | – | – | – | – | ✓ stream | ✓ spill |
| **Best for** | models, complex R objects | fast model/list caching | large df with partial reads | CSV exchange with non-R users | cross-language fast exchange | big data, cloud analytics | SQL workflows, out-of-memory |

: Format comparison across key properties {#tbl-formats}

## Read More

**Package documentation**

- `saveRDS` / `readRDS`: [base R serialization reference](https://rdrr.io/r/base/readRDS.html)
- **qs2**: [CRAN](https://CRAN.R-project.org/package=qs2) · [GitHub](https://github.com/traversc/qs)
- **fst**: [fstpackage.org](https://www.fstpackage.org/) · [CRAN](https://CRAN.R-project.org/package=fst)
- **data.table** `fwrite`/`fread`: [rdatatable.gitlab.io/data.table](https://rdatatable.gitlab.io/data.table/)
- **Arrow for R**: [arrow.apache.org/docs/r](https://arrow.apache.org/docs/r/) — includes vignettes on datasets, S3, and Parquet
- Apache Parquet format: [parquet.apache.org](https://parquet.apache.org/)
- **DuckDB R API**: [duckdb.org/docs/api/r](https://duckdb.org/docs/api/r) · **duckplyr** (drop-in dplyr backend): [duckplyr.tidyverse.org](https://duckplyr.tidyverse.org/)

**Benchmarks**

- fastverse benchmarks wiki: <https://github.com/fastverse/fastverse/wiki/Benchmarks> — curated index of collapse, data.table, and Arrow comparisons
- DuckDB Labs db-benchmark (group-by and join at 0.5 GB, 5 GB, 50 GB): <https://duckdblabs.github.io/db-benchmark/>
- Adrian Antico multi-tool benchmarks (1M–1B rows, 8 operations): <https://github.com/AdrianAntico/Benchmarks#background>

**Background reading**

- Burns, P. (2011). *The R Inferno*. [burns-stat.com](https://www.burns-stat.com/pages/Tutor/R_inferno.pdf) — § 8 covers numeric storage and write/read pitfalls that motivate binary serialization.
